Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-10-25 manuscript/main.tex
final main text edits

% Limit 350 words
\begin{abstract}
Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses. This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability. Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (USA-NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites. We built statistical and process based phenology models for species common to each dataset. From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both USA-NPN and LTER data. We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased. Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen. Predictions for USA-NPN data had the lowest error when using models built from the USA-NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built. This difference is likely due to variation in phenological requirements within species. Models using the USA-NPN dataset can integrate parameters over a large spatial scale while those using an LTER dataset can only estimate parameters for a single location. Accordingly, the choice of dataset depends on the research question. Inferences about species-specific phenological requirements are best made with LTER data, and if USA-NPN or similar data are all that is available, then analyses should be limited to simple models. Large-scale predictive modeling is best done with the larger-scale USA-NPN data, which has high spatial representation and a large regional species pool. LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well. Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
\end{abstract}

@@

The USA National Phenology Network (USA-NPN) protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the USA-NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). Three of the LTER datasets (Hubbard Brook, Harvard Forest, and H.J. Andrews) had a sampling frequency of 3-7 days during the growing season. The Jornada dataset had a sampling frequency of 30 days. As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1). Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species.

@@

Finally, we directly evaluated performance using out-of-sample errors from the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations. Using the RMSE values from held out observations, we compared the performance of LTER and USA-NPN derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better. We used a t-test to test the difference from zero in these values. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values as well as Pearson correlation coefficients are provided in the supplement for specific species (Fig. S5-S7) and with all observations aggregated together (Table S2).

@@

Throughout the analysis there were no qualitative differences between a 30-day or 15-day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset. Results presented here reflect the 30 day cutoff, see the supplementary figures S2-S4 for matching figures using a 15 day cutoff.

The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$). The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig. 2). The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig. 2). The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}. While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig. S10-S11).

When comparing estimates of phenological events between the two sets of models, many USA-NPN and LTER models produced similar estimates (Fig. 3). The \textit{Fixed GDD model} had the highest correlation between the two models sets at USA-NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively). Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}. The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at USA-NPN sites. The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) improved the correlation over the \textit{GDD model} only slightly at LTER sites (from 0.51 to 0.52) and did not improve the correlation at USA-NPN sites.

When comparing the predictive performance using out-of-sample errors, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig. 4). The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations. Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall. The \textit{GDD} and \textit{Uniforc} models made the best out of sample predictions, having the lowest RMSE and Pearson coefficient when aggregating all observations together (Table S2). One exception was that the Fixed GDD model had a slightly higher Pearson value when using LTER-derived models to make predictions for USA-NPN data. The best model for specific species and phenophases varied, but was commonly the \textit{Uniforc} and \textit{GDD} models (Figs. S5-S6).

@@

Data used to build phenology models typically falls into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., USA-NPN data) (Table 3). This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and makes it necessary to understand how to choose between, or combine, data sources \citep{hanks2018}. We explored this issue for phenology modeling in relation to making predictions and inferring process from models. For inference we found that models based on different data sources resulted in different parameter estimates for all but the simplest models. For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out-of-sample data from the data type to which they were fit. These results are consistent with other research showing that phenology model performance decreases when transferring single-site models to other locations \citep{garcia-mozo2008, xu2013, basler2016}, and with the call for models that better incorporate spatial variation in phenology requirements \citep{richardson2013, chuine2017}. Understanding and making predictions for the phenology of a single location is best served by intensive local-scale data, when available, but large-scale datasets work better for extrapolating phenology predictions across a species range. Thus, the best choice of both data and models depends on the desired research goals.

In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the \textit{Fixed GDD} (Fig. 2). These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters. First, the differences could result from limitations in the sampling of the USA-NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates. If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the \textit{Fixed GDD model} is best for making inferences when USA-NPN data are all that is available. Second, spatial variation (e.g. from local adaptation, acclimation, microclimates, or plant age)  in phenology requirements and drivers could contribute to these differences \citep{diez2012, zhang2017}. Models built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site. In this case, USA-NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site. The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary. Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same. Information about which of these issues may be causing the differences between datasets can be explored using the analyses in the current study, as will be explained below.

Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two \textit{GDD} models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig. S10-S11). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S8-S9). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates that were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions. Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Fig. S5 \& S6). This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}. In combination with our results showing observed differences in parameter estimates between LTER sites (Fig. S8-S9), this suggests that variation in phenological requirements across the range is likely important. However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses. Specifically, correspondence between parameter estimates (Fig. 2), estimates of phenological events (Fig. 3), and out-of-sample error rates (Fig. 4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively. This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the USA-NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by having a 40-year time-series across a large region. \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research. Large-scale phenology datasets, like USA-NPN, will be key in addressing this and other phenological research needs.

In addition to exploring differences between phenology datasets, our analyses provide guidance on which models to use when making predictions at a local scale using models built from large-scale data, or vice versa. Among the eight models tested, the \textit{Uniforc} and \textit{GDD} models performed the best overall in the cross dataset comparison in terms of Pearson correlation and RMSE (Fig. S5-S6, Table S2). The \textit{GDD} model has one less parameter than the \textit{Uniforc} model, thus the \textit{GDD} model is a suitable choice for making predictions when there is little to no information at the location of interest (e.g. making phenology forecasts at a new location distant from any observed data). This guidance can vary between species though and model testing should still be performed when suitable data is available.

@@

@@

\begin{adjustbox}{angle=90}
%%renewcommand{\arraystretch}{2}\tabcolsep=5pt
    \begin{tabular}{ | l | l | l | l |}
    \hline
    Dataset Name & Habitat &  Phenological Event (Num. Species) & Reference \\ \hline
    Harvard Forest & N.E. Deciduous Forest & Budburst (17) Flowering (7) & \citep{okeefe2015} \\
    Jornada Experimental Range & Chihuahuan Desert & Flowering (2) &  \\
    H.J. Andrews Experimental Forest & N.W. Wet Coniferous Forest & Budburst (5) Flowering (4) & \citep{schulze2017} \\
    Hubbard Brook & N.E. Deciduous Forest & Budburst (3) & \citep{bailey2018} \\
    \hline
    \end{tabular}
\end{adjustbox} \newline
\textbf{Table 1}: LTER datasets used in the analysis

\newpage
\begin{adjustbox}{angle=90}
\small
%%{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T_{base}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T_{base}, 5) $  & 4 & \citep{blumel2012} \\
    Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
    MSB & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
    \hline
    \end{tabular}
\end{adjustbox} \newline 
\textbf{Table 2}: Phenology models used in the analysis

@@

%%%%%%%%%%%%%%%%%%%%%%
%% Figure 1, the map
%%%%%%%%%%%%%%%%%%%%%%

>>>>

% Limit 350 words
\begin{abstract}
Large-scale observational data from citizen science efforts are becoming increasingly common in ecology, and researchers often choose between these and data from intensive local-scale studies for their analyses. This choice has potential trade-offs related to spatial scale, observer variance, and inter-annual variability. Here we explored this issue with phenology by comparing models built using data from the large-scale, citizen science USA National Phenology Network (USA-NPN) effort with models built using data from more intensive studies at Long Term Ecological Research (LTER) sites. We built statistical and process based phenology models for species common to each dataset. From these models we compared parameter estimates, estimates of phenological events, and out-of-sample errors between models derived from both USA-NPN and LTER data. We found that model parameter estimates for the same species were most similar between the two datasets when using simple models, but parameter estimates varied widely as model complexity increased. Despite this, estimates for the date of phenological events and out-of-sample errors were similar, regardless of the model chosen. Predictions for USA-NPN data had the lowest error when using models built from the USA-NPN data, while LTER predictions were best made using LTER-derived models, confirming that models perform best when applied at the same scale they were built. This difference in the cross-scale model comparison is likely due to variation in phenological requirements within species. Models using the USA-NPN dataset can integrate parameters over a large spatial scale while those using an LTER dataset can only estimate parameters for a single location. Accordingly, the choice of dataset depends on the research question. Inferences about species-specific phenological requirements are best made with LTER data, and if USA-NPN or similar data are all that is available, then analyses should be limited to simple models. Large-scale predictive modeling is best done with the larger-scale USA-NPN data, which has high spatial representation and a large regional species pool. LTER datasets, on the other hand, have high site fidelity and thus characterize inter-annual variability extremely well. Future research aimed at forecasting phenology events for particular species over larger scales should develop models which integrate the strengths of both datasets.
\end{abstract}

@@

The USA National Phenology Network (USA-NPN) protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the USA-NPN datasets are located across the U.S. and generally clustered around populated areas (Figure 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Figure 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

@@

In the LTER datasets observation metrics varied widely due to different protocols. To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see Appendix S1 for details). Three of the LTER datasets (Hubbard Brook, Harvard Forest, and H.J. Andrews) had a sampling frequency of 3-7 days during the growing season. The Jornada dataset had a sampling frequency of 30 days. As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& Appendix S2: Table S1). Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species.

@@

Finally, we directly evaluated performance using out-of-sample errors from the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations. Using the RMSE values from held out observations, we compared the performance of LTER and USA-NPN derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better. We used a t-test to test the difference from zero in these values. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values as well as Pearson correlation coefficients are provided in the supplement for specific species (Appendix S2: Figure S5-S7) and with all observations aggregated together (Appendix S2: Table S2).

@@

Throughout the analysis there were no qualitative differences between a 30-day or 15-day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset. Results presented here reflect the 30 day cutoff; see the supplementary figures S2-S4 in Appendix S2 for matching figures using a 15 day cutoff.

The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$). The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Figure 2). The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Figure 2). The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}. While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Appendix S2: Figure S10-S11).

When comparing estimates of phenological events between the two sets of models, many USA-NPN and LTER models produced similar estimates (Figure 3). The \textit{Fixed GDD model} had the highest correlation between the two model sets at USA-NPN sites ($R^2 = 0.82$), while the \textit{GDD}, \textit{M1}, and \textit{Uniforc} models had the highest correlation at LTER sites ($R^2 =$ 0.51, 0.52, and 0.51, respectively). Comparing models with spatial corrections to the non-spatial alternatives, the \textit{MSB} (an extension of the \textit{Alternating model} with a spatial correction based on mean spring temperature, see Table 2 and Methods) improved the correlation between the two datasets over the \textit{Alternating model}. The \textit{MSB model} improved the $R^2$ from 0.36 to 0.45 at LTER sites, and from -0.23 to -0.15 at USA-NPN sites. The \textit{M1 model} (an extension of the \textit{GDD model} with a spatial correction based on day length) improved the correlation over the \textit{GDD model} only slightly at LTER sites (from 0.51 to 0.52) and did not improve the correlation at USA-NPN sites.

When comparing the predictive performance using out-of-sample errors, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Figure 4). The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations. Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall. The \textit{GDD} and \textit{Uniforc} models made the best out of sample predictions, having the lowest RMSE and Pearson coefficient when aggregating all observations together (Appendix S2: Table S2). One exception was that the Fixed GDD model had a slightly higher Pearson value when using LTER-derived models to make predictions for USA-NPN data. The best model for specific species and phenophases varied, but was commonly the \textit{Uniforc} or \textit{GDD} models (Appendix S2: Figs. S5-S6).

@@

Data used to build phenology models typically fall into two categories: intensive long-term data with long time-series at a small number of locations (e.g., LTER data in this study), and large-scale data with less intensive sampling at hundreds of locations (e.g., USA-NPN data) (Table 3). This data scenario--a small amount of intensive data and a large amount of less intensive data--is common in many areas of science and makes it necessary to understand how to choose between, or combine, data sources \citep{hanks2018}. We explored this issue for phenology modeling in relation to making predictions and inferring process from models. For inference we found that models based on different data sources resulted in different parameter estimates for all but the simplest models. For prediction we found that models fit to different data sources tended to make similar predictions, but that models better predicted out-of-sample data from the data type to which they were fit. These results are consistent with other research showing that phenology model performance decreases when transferring single-site models to other locations \citep{garcia-mozo2008, xu2013, basler2016}, and with the call for models that better incorporate spatial variation in phenology requirements \citep{richardson2013, chuine2017}. Understanding and making predictions for the phenology of a single location is best served by intensive local-scale data, when available, but large-scale datasets work better for extrapolating phenology predictions across a species range. Thus, the best choice of both data and models depends on the desired research goals.

In this study, parameter estimates differed widely within the same phenology model when fit to the two different types of data, except for the simplest process-oriented model: the \textit{Fixed GDD} (Figure 2). These differences may be caused by a variety of factors that have different implications for interpreting process-oriented models and their parameters. First, the differences could result from limitations in the sampling of the USA-NPN dataset, such as irregular sampling of the same location within or between seasons, leading to less accurate parameter estimates. If this is the case, it would suggest that using LTER data is ideal for making inferences about plant physiology, and that focusing on the \textit{Fixed GDD model} is best for making inferences when USA-NPN data are all that is available. Second, spatial variation (e.g. from local adaptation, acclimation, microclimates, or plant age)  in phenology requirements and drivers could contribute to these differences \citep{diez2012, zhang2017}. Models built using USA-NPN data integrate over that spatial variation, while models built using LTER data only estimate the phenological requirements for a specific site. In this case, USA-NPN data would provide a better estimate of the general phenological requirements of a species, but LTER data would provide a more accurate understanding for a single site. The best solution to this issue would be the development of models that accurately incorporate spatial variation, such as including genetic variation between different populations \citep{chuine2017}, although localized models could also be generated when large-scale predictions are unnecessary. Third, these differences could result from issues with model identifiability: since different parameter values can yield nearly identical estimates of phenological events, parameter estimates can differ between datasets even when the underlying processes generating the data are the same. Information about which of these issues may be causing the differences between datasets can be explored using the analyses in the current study, as will be explained below.

Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Figure 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two \textit{GDD} models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Appendix S2: Figure S10-S11). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Appendix S2: Figure S8-S9). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates that were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions. Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Appendix S2: Figure S5 \& S6). This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}. In combination with our results showing observed differences in parameter estimates between LTER sites (Appendix S2: Figure S8-S9), this suggests that variation in phenological requirements across the range is likely important. However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses. Specifically, correspondence between parameter estimates (Figure 2), estimates of phenological events (Figure 3), and out-of-sample error rates (Figure 4) for the \textit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively. This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the USA-NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by having a 40-year time-series across a large region. \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research. Large-scale phenology datasets, like USA-NPN, will be key in addressing this and other phenological research needs.

In addition to exploring differences between phenology datasets, our analyses provide guidance on which models to use when making predictions at a local scale using models built from large-scale data, or vice versa. Among the eight models tested, the \textit{Uniforc} and \textit{GDD} models performed the best overall in the cross dataset comparison in terms of Pearson correlation and RMSE (Appendix S2: Figure S5-S6, Table S2). The \textit{GDD} model has one less parameter than the \textit{Uniforc} model, thus the \textit{GDD} model is a suitable choice for making predictions when there is little to no information at the location of interest (e.g. making phenology forecasts at a new location distant from any observed data). This guidance can vary between species, though, and model testing should still be performed when suitable data are available.

@@

\newpage
\scriptsize
%%renewcommand{\arraystretch}{2}\tabcolsep=5pt
    \begin{tabular}{ | l | l | l | l |}
    \hline
    Dataset Name & Habitat &  Phenological Event (Num. Species) & Reference \\ \hline
    Harvard Forest & N.E. Deciduous Forest & Budburst (17) Flowering (7) & \citep{okeefe2015} \\
    Jornada Experimental Range & Chihuahuan Desert & Flowering (2) &  \\
    H.J. Andrews Experimental Forest & N.W. Wet Coniferous Forest & Budburst (5) Flowering (4) & \citep{schulze2017} \\
    Hubbard Brook & N.E. Deciduous Forest & Budburst (3) & \citep{bailey2018} \\
    \hline
    \end{tabular}
\newline
\textbf{Table 1}: LTER datasets used in the analysis

@@

\scriptsize
%%{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T_{base}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
    Fixed GDD &$\sum_{t=1}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
    Alternating & $\sum_{t=1}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T_{base}, 5) $  & 4 & \citep{blumel2012} \\
    MSB & $\sum_{t=1}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
    \hline
    \end{tabular}
\newline

\textbf{Table 2}: Phenology models used in the analysis. For all models, except the Naive and Linear models, the daily mean temperature $T_{i}$ is first transformed via the specified forcing equation. The cumulative sum of forcing is then calculated from a specific start date (either $DOY=1$ or using the fitted parameter $t_{1}$). The phenological event is estimated as the $DOY$ in which cumulative forcing is greater than or equal to the specified total required forcing (either $F^{*}$ or the specified equation). Parameters for each model are as follows: For the Naive model $\overline{DOY}$ is the mean day of year (ie. the Julian date) of a phenological event; for the Linear model $\beta_{1}$ and $\beta_{2}$ are the intercept and slope, respectively and $T_{mean}$ is the average daily temperature between January 1 and March 31; for the GDD model $F^{*}$ is the total accumulated forcing required, $t_{1}$ is the start date of forcing accumulation, and $T_{base}$ is the threshold daily mean temperature above which forcing accumulates; for the Fixed GDD model $F^{*}$ is the total accumulated forcing required; for the Alternating model $NCD$ is the number of chill days (daily mean temperature below 0$^{\circ}$C) from $DOY=0$ to the $DOY$ of the phenological event, $a$, $b$, and $c$ are the three fitted model coefficients; for the Uniforc model, is $F^{*}$ is the total accumulated forcing required, $t_{1}$ is the start date of forcing accumulation, and $b$ and $c$ are two additional fitted parameters which define the sigmoid function; the M1 model is the same as the GDD model, but with the additional fitted parameter $k$ that adjusts the total forcing accumulation according to day length; the MSB model is the same as the Alternating model, but with the additional fitted parameter $d$ to correct the model according to mean spring temperature.

@@

%%%%%%%%%%%%%%%%%%%%%%
%% Figure 1 Legend, the map
%%%%%%%%%%%%%%%%%%%%%%
